---
title: "Lexical Foundations and Vocabulary Construction in NLP"
author: "Natali Angarita Escolar — Universidad del Norte"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
subtitle: 'Applied Activity: *Somewhere Over the Rainbow*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(42)
```

```{r libraries, message=FALSE, warning=FALSE}
# ── Required libraries ──────────────────────────────────────────────────────
library(tidytext)
library(dplyr)
library(stringr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(SnowballC)
library(stopwords)
library(tidyr)
library(tibble)
library(tinytex)
options(knitr.table.format = "latex")

# ── Library versions ─────────────────────────────────────────────────────────
pkg_versions <- data.frame(
  Package = c("tidytext","dplyr","stringr","knitr",
              "kableExtra","ggplot2","SnowballC","stopwords","tidyr","tibble"),
  Version = sapply(c("tidytext","dplyr","stringr","knitr",
                     "kableExtra","ggplot2","SnowballC","stopwords","tidyr","tibble"),
                   function(p) as.character(packageVersion(p)))
)

kable(pkg_versions, caption = "Table 0.1: Library versions used in this document") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

---

# Text Description

**"Somewhere Over the Rainbow"** is one of the most iconic songs in the history of American popular
music. It was composed by **Harold Arlen** with lyrics by **E.Y. Harburg**, and was introduced by
**Judy Garland** in the 1939 film *The Wizard of Oz*. The song is widely considered part of the
**public domain** for lyrical study and educational purposes, and has been recorded by hundreds of
artists, most famously by Israel Kamakawiwoʻole in 1993.

The song expresses a **longing for an idealized world** beyond hardship — a place where dreams
come true, troubles vanish, and the impossible becomes possible. Its language is rich in
**emotional vocabulary**, **nature imagery**, **wishful modality** (*wish*, *dare*, *find*),
and **spatial metaphors** (*over*, *above*, *beyond*), making it an ideal text for NLP analysis.

**Fragment selected for analysis:**

> *Someday I'll wish upon a star*
> *Wake up where the clouds are far behind me*
> *Where trouble melts like lemon drops*
> *High above the chimney tops, that's where*
> *You'll find me*
> *Oh, somewhere over the rainbow*
> *Blue birds fly*
> *And the dream that you dare to*
> *Oh, why, oh, why can't I?*

```{r define-text}
# ── Song fragment: one element per line ──────────────────────────────────────
song_lines <- c(
  "Someday I'll wish upon a star",
  "Wake up where the clouds are far behind me",
  "Where trouble melts like lemon drops",
  "High above the chimney tops that's where",
  "You'll find me",
  "Oh somewhere over the rainbow",
  "Blue birds fly",
  "And the dream that you dare to",
  "Oh why oh why can't I"
)

# ── Full text as a single string ─────────────────────────────────────────────
song_text <- paste(song_lines, collapse = " ")
cat("Full text:\n", song_text, "\n")
```

---

# Lexicons

A **lexicon** is the set of words recognized and used within a language or domain. In NLP it
functions as a dictionary that defines which terms are meaningful in a given context. Words outside
the lexicon are typically ignored, replaced, or decomposed into smaller units
(LLinás Solano, 2026, §3).

The table below presents a lexicon of **15 entries** derived directly from the song fragment.
Each entry includes the lexical item, a conceptual category, and a short interpretation.

```{r lexicon-table}
# ── Lexicon: 15 entries ──────────────────────────────────────────────────────
lexicon <- tribble(
  ~`Lexical Item`, ~`Category`,          ~`Description / Interpretation`,
  "someday",       "Temporal adverb",    "Refers to an unspecified future moment; expresses hope",
  "wish",          "Action / Desire",    "Verb expressing a longing for something not yet real",
  "star",          "Nature / Symbol",    "Celestial body; culturally symbolizes wishes and dreams",
  "wake",          "Action",             "To come out of sleep; signals a transition or awakening",
  "clouds",        "Nature",             "Atmospheric formations; metaphor for obstacles or distance",
  "far",           "Spatial adjective",  "At a great distance; emphasizes separation from trouble",
  "trouble",       "Emotion / State",    "Difficulty or distress; the negative reality being escaped",
  "melts",         "Action / Metaphor",  "To dissolve; implies troubles disappearing like solids",
  "lemon",         "Nature / Sensory",   "Citrus fruit used metaphorically for bitterness turned sweet",
  "drops",         "Nature / Action",    "Small falling quantities; part of the compound 'lemon drops'",
  "chimney",       "Place / Object",     "Part of a rooftop; grounds the song in a domestic setting",
  "rainbow",       "Nature / Symbol",    "Optical phenomenon; central symbol of hope and aspiration",
  "blue",          "Color / Emotion",    "Color associated with sky and birds; also connotes longing",
  "dream",         "Emotion / Concept",  "An aspiration or vision; core thematic word of the song",
  "dare",          "Action / Emotion",   "To have the courage to attempt; implies risk in hoping"
)

kable(lexicon,
      caption = "Table 3.1: Song lexicon — 15 entries derived from the fragment") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE) %>%
  column_spec(1, bold = TRUE)
```

---

# Phonemes, Graphemes, and Morphemes

Before building a vocabulary, it is useful to understand three basic linguistic units
(LLinás Solano, 2026, §4):

- **Phonemes**: the smallest *sound* units that distinguish meaning in spoken language.
- **Graphemes**: letters or letter groups that *represent* phonemes in written language.
- **Morphemes**: the smallest units that *carry meaning* (roots, prefixes, suffixes).

The table below illustrates all three concepts using words taken directly from the song fragment.

```{r pgm-table}
steelblue <- "#4682B4"
# ── Phonemes, graphemes, morphemes illustration ──────────────────────────────
pgm <- tribble(
  ~Word,       ~Phonemes,                          ~Graphemes,                        ~Morphemes,
  "rainbow",   "/r/ /eɪ/ /n/ /b/ /oʊ/  — 5 sounds","r, ai, n, b, ow  — 5 graphemes", "rain + bow  — 2 morphemes",
  "somewhere", "/s/ /ʌ/ /m/ /w/ /ɛ/ /r/  — 6 sounds","s, o, m, e, w, h, e, r, e — 9 graphemes","some + where  — 2 morphemes",
  "bluebirds", "/b/ /l/ /uː/ /b/ /ɜːr/ /d/ /z/  — 7","b, l, ue, b, ir, d, s — 7 graphemes",  "blue + bird + s  — 3 morphemes",
  "trouble",   "/t/ /r/ /ʌ/ /b/ /əl/  — 5 sounds",   "t, r, ou, b, l, e  — 6 graphemes",    "trouble  — 1 morpheme (root)",
  "chimney",   "/tʃ/ /ɪ/ /m/ /n/ /iː/  — 5 sounds",  "ch, i, m, n, ey  — 5 graphemes",      "chimney  — 1 morpheme (root)",
  "someday",   "/s/ /ʌ/ /m/ /d/ /eɪ/  — 5 sounds",   "s, o, m, e, d, a, y  — 7 graphemes",  "some + day  — 2 morphemes"
)

kable(pgm,
      caption = "Table 4.1: Phonemes, graphemes and morphemes illustrated with song words") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE) %>%
  column_spec(1, bold = TRUE, color = steelblue)
```

**Key distinctions:**

- *rainbow* has **5 phonemes** but the grapheme `ow` represents a single sound `/oʊ/`,
  illustrating that phonemes ≠ letters.
- *bluebirds* decomposes into **3 morphemes** (`blue` + `bird` + `s`), showing how meaning
  accumulates through morphological structure.
- *somewhere* and *someday* both share the prefix morpheme `some-`, demonstrating
  productive morphology.

---

# Tokenization

Tokenization is the process of segmenting text into smaller units called **tokens**
(LLinás Solano, 2026, §5). We apply and compare three strategies below.

## Sentence Tokenization

```{r sentence-tokenization}
# ── Sentence tokenization: each line = one sentence ──────────────────────────
sentence_tokens <- song_lines

n_sentences <- length(sentence_tokens)
cat("Total number of sentence tokens:", n_sentences, "\n\n")

sent_df <- data.frame(
  `Sentence #` = seq_along(sentence_tokens),
  `Sentence token` = sentence_tokens,
  check.names = FALSE
)

kable(sent_df,
      caption = "Table 5.1: Sentence-level tokenization") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE)
```

## Word Tokenization

```{r word-tokenization}
# ── Word tokenization ─────────────────────────────────────────────────────────
# Remove punctuation and split on whitespace
word_tokens <- unlist(str_split(
  str_replace_all(song_text, "[[:punct:]]", ""),
  "\\s+"
))
word_tokens <- word_tokens[word_tokens != ""]

n_words <- length(word_tokens)
cat("Total number of word tokens:", n_words, "\n\n")

# Frequency table
word_freq <- sort(table(tolower(word_tokens)), decreasing = TRUE)
word_freq_df <- data.frame(
  Token     = names(word_freq),
  Frequency = as.integer(word_freq)
)

kable(head(word_freq_df, 15),
      caption = "Table 5.2: Top-15 most frequent word tokens") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

```{r word-token-plot, fig.cap="Figure 5.1: Top-10 most frequent word tokens"}
ggplot(head(word_freq_df, 10),
       aes(x = reorder(Token, Frequency), y = Frequency)) +
  geom_col(fill = "steelblue", width = 0.7) +
  coord_flip() +
  labs(title = "Top-10 Word Tokens - Somewhere Over the Rainbow",
       x = "Token", y = "Frequency") +
  theme_minimal(base_size = 13)
```

## Character-Level Tokenization

```{r char-tokenization}
# ── Character-level tokenization ─────────────────────────────────────────────
clean_text   <- str_replace_all(song_text, "[[:punct:]\\s]", "")
char_tokens  <- unlist(str_split(tolower(clean_text), ""))

n_chars <- length(char_tokens)
cat("Total number of character tokens:", n_chars, "\n\n")

char_freq <- sort(table(char_tokens), decreasing = TRUE)
char_freq_df <- data.frame(
  Character = names(char_freq),
  Frequency = as.integer(char_freq)
)

kable(head(char_freq_df, 10),
      caption = "Table 5.3: Top-10 most frequent character tokens") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

## Summary of Tokenization Results

```{r tokenization-summary}
tok_summary <- data.frame(
  Strategy            = c("Sentence tokenization",
                           "Word tokenization",
                           "Character tokenization"),
  `Total tokens`      = c(n_sentences, n_words, n_chars),
  `Most frequent`     = c(
    "All lines appear once",
    paste(head(word_freq_df$Token, 3), collapse = ", "),
    paste(head(char_freq_df$Character, 3), collapse = ", ")
  ),
  check.names = FALSE
)

kable(tok_summary,
      caption = "Table 5.4: Comparison of tokenization strategies") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE)
```

**Interpretation:** Word tokenization reveals that function words (*oh*, *where*, *the*) dominate
by frequency, which is expected in short lyrical texts. Character tokenization shows vowels (`e`,
`o`, `a`) as the most common characters — consistent with English phonology. The 9-sentence
structure reflects the verse-by-verse organization of the fragment.

---

# Different Types of Tokenizers

Following LLinás Solano (2026, §6), we apply **rule-based tokenization** using regular
expressions (`stringr`) to extract specific entities from the song text.
We design **three** regex patterns targeting different linguistic structures.

```{r regex-setup}
# Full text with original punctuation kept for regex demonstration
song_text_raw <- paste(c(
  "Someday I'll wish upon a star",
  "Wake up where the clouds are far behind me",
  "Where trouble melts like lemon drops",
  "High above the chimney tops, that's where",
  "You'll find me",
  "Oh, somewhere over the rainbow",
  "Blue birds fly",
  "And the dream that you dare to",
  "Oh, why, oh, why can't I?"
), collapse = " ")

cat("Raw text used for regex:\n", song_text_raw, "\n")
```

## Regex 1 — Contractions (words with apostrophes)

```{r regex1}
# Pattern: word characters + apostrophe + word characters
pattern1  <- "[A-Za-z]+\\'[A-Za-z]+"
matches1  <- str_extract_all(song_text_raw, pattern1)[[1]]

cat("Pattern:", pattern1, "\n")
cat("Matches found:", length(matches1), "\n")

res1 <- data.frame(
  Match = matches1,
  Explanation = c(
    "I'll  → I + will  (future contraction)",
    "that's → that + is (copula contraction)",
    "You'll → You + will (future contraction)",
    "can't  → can + not (negation contraction)"
  )[seq_along(matches1)]
)

kable(res1,
      caption = "Table 6.1: Regex 1 — Contractions extracted from the text") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

**What this pattern captures:** Sequences of letters separated by an apostrophe (`'`).
This is directly relevant to the tokenization challenge discussed in §5 of the course notes:
simple `split()` treats *I'll* as one token, while linguistically it represents two morphemes.

## Regex 2 — Capitalised words (potential proper nouns / sentence starters)

```{r regex2}
# Pattern: word starting with an uppercase letter
pattern2 <- "\\b[A-Z][a-z]+"
matches2 <- str_extract_all(song_text_raw, pattern2)[[1]]

cat("Pattern:", pattern2, "\n")
cat("Matches found:", length(matches2), "\n")

res2 <- data.frame(
  `Capitalised token` = matches2,
  check.names = FALSE
)

kable(res2,
      caption = "Table 6.2: Regex 2 — Capitalised words extracted") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

**What this pattern captures:** Words beginning with an uppercase letter (`[A-Z]`) followed by
one or more lowercase letters. This identifies sentence-initial words and potential named entities.
In this text, all matches are sentence starters — there are no proper nouns, which reflects the
abstract, universal nature of the lyrics.

## Regex 3 — Monosyllabic exclamations / interjections

```{r regex3}
# Pattern: "Oh" or "why" (interjections repeated in the fragment)
pattern3 <- "\\b(Oh|oh|why|Why)\\b"
matches3 <- str_extract_all(song_text_raw, pattern3)[[1]]

cat("Pattern:", pattern3, "\n")
cat("Matches found:", length(matches3), "\n")

res3 <- data.frame(
  Match = matches3,
  Type  = ifelse(tolower(matches3) == "oh", "Exclamation / filler", "Interrogative adverb")
)

kable(res3,
      caption = "Table 6.3: Regex 3 — Interjections and interrogatives") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

**What this pattern captures:** The alternation `(Oh|oh|why|Why)` with word boundaries (`\\b`)
extracts the emotional interjections that give the closing lines their rhetorical force.
As noted in the course (§6.0.1), `a|b` in regex matches either option — here applied to
case-insensitive variants of the same word.

---

# Word Normalization

Word normalization maps different surface forms to a consistent representation, controlling
vocabulary size and reducing noise (LLinás Solano, 2026, §7). We apply the following pipeline.

```{r normalization-pipeline}
# ── Step 0: Original tokens (before normalization) ───────────────────────────
original_tokens <- unlist(str_split(song_text_raw, "\\s+"))
original_tokens <- original_tokens[original_tokens != ""]

cat("Original token count:", length(original_tokens), "\n")
cat("Original vocabulary size:", length(unique(tolower(original_tokens))), "\n\n")
```

## Step 1 — Lowercasing

```{r lowercasing}
tokens_lower <- tolower(original_tokens)
cat("After lowercasing — unique forms:", length(unique(tokens_lower)), "\n")
cat("Sample:", paste(head(tokens_lower, 10), collapse = " "), "\n")
```

## Step 2 — Punctuation Removal

```{r punct-removal}
tokens_nopunct <- str_replace_all(tokens_lower, "[[:punct:]]", "")
tokens_nopunct <- tokens_nopunct[tokens_nopunct != ""]

cat("After punctuation removal — tokens:", length(tokens_nopunct), "\n")
cat("Vocabulary size:", length(unique(tokens_nopunct)), "\n")
cat("Sample:", paste(head(tokens_nopunct, 10), collapse = " "), "\n")
```

## Step 3 — Stopword Removal

```{r stopword-removal}
en_stopwords <- stopwords("en", source = "snowball")

tokens_nostop <- tokens_nopunct[!tokens_nopunct %in% en_stopwords]

cat("After stopword removal — tokens:", length(tokens_nostop), "\n")
cat("Vocabulary size:", length(unique(tokens_nostop)), "\n")
cat("Remaining tokens:", paste(tokens_nostop, collapse = " "), "\n")
```

## Step 4 — Stemming

```{r stemming}
tokens_stemmed <- wordStem(tokens_nostop, language = "english")

cat("After stemming — unique stems:", length(unique(tokens_stemmed)), "\n")
cat("Stemmed tokens:", paste(tokens_stemmed, collapse = " "), "\n")
```

## Vocabulary Comparison: Before vs After Normalization

```{r vocab-comparison}
# Vocabulary at each stage
vocab_original <- unique(tolower(original_tokens))
vocab_nopunct  <- unique(tokens_nopunct)
vocab_nostop   <- unique(tokens_nostop)
vocab_stemmed  <- unique(tokens_stemmed)

# Comparison table
vocab_comp <- data.frame(
  Stage = c(
    "0. Original (raw)",
    "1. Lowercasing",
    "2. Punctuation removal",
    "3. Stopword removal",
    "4. Stemming"
  ),
  `Token count` = c(
    length(original_tokens),
    length(tokens_lower),
    length(tokens_nopunct),
    length(tokens_nostop),
    length(tokens_stemmed)
  ),
  `Vocabulary size` = c(
    length(vocab_original),
    length(unique(tokens_lower)),
    length(vocab_nopunct),
    length(vocab_nostop),
    length(vocab_stemmed)
  ),
  `Example tokens` = c(
    paste(head(original_tokens, 5), collapse = ", "),
    paste(head(tokens_lower,    5), collapse = ", "),
    paste(head(tokens_nopunct,  5), collapse = ", "),
    paste(head(tokens_nostop,   5), collapse = ", "),
    paste(head(tokens_stemmed,  5), collapse = ", ")
  ),
  check.names = FALSE
)

kable(vocab_comp,
      caption = "Table 7.1: Vocabulary before and after each normalization step") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE) %>%
  row_spec(5, bold = TRUE, background = "#e8f4fd")
```

```{r vocab-plot, fig.cap="Figure 7.1: Vocabulary size reduction across normalization stages"}
ggplot(vocab_comp,
       aes(x = Stage, y = `Vocabulary size`, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(color = "steelblue", size = 4) +
  geom_text(aes(label = `Vocabulary size`), vjust = -0.8, size = 4) +
  labs(title = "Vocabulary Size Reduction Through Normalization",
       subtitle = "Somewhere Over the Rainbow - NLP Pipeline",
       x = NULL, y = "Unique vocabulary items") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

**Discussion of observed changes:**

- **Lowercasing** collapses case variants (e.g., *Oh* and *oh*) into a single type, immediately
  reducing vocabulary redundancy.
- **Punctuation removal** eliminates non-lexical characters and also resolves contractions like
  *I'll* and *can't* into single tokens without apostrophes (`ill`, `cant`) — a simplification
  with semantic costs.
- **Stopword removal** has the largest impact: function words (*oh*, *the*, *where*, *a*, *to*,
  *I*) dominate the lyrics but carry little standalone semantic weight. Removing them exposes the
  content words: *someday*, *wish*, *star*, *trouble*, *rainbow*, *dream*.
- **Stemming** further compresses the vocabulary by mapping inflected forms to their root
  (e.g., *melts* → *melt*, *birds* → *bird*, *drops* → *drop*), reducing the final vocabulary
  to its most essential lexical cores.

---

# Summary

Lexical choices, tokenization, and normalization are three interrelated stages that fundamentally
determine how a text is represented in an NLP system. In *Somewhere Over the Rainbow*, the
emotional power of the lyrics relies on carefully chosen words — *wish*, *dream*, *dare* — that
belong to specific conceptual categories (desire, aspiration, uncertainty). Word tokenization
reveals that function words dominate raw frequency counts, while content words carry the
thematic weight of the text. Applying normalization progressively — lowercasing, punctuation
removal, stopword filtering, and stemming — reduced the vocabulary from its original size to a
compact set of meaningful stems, showing how preprocessing decisions shape what a model
"sees." Crucially, each step involves a trade-off: removing stopwords like *why* or *oh* cleans
noise but also erases the rhetorical texture of the verse. As LLinás Solano (2026) notes, these
choices are modeling decisions that must be aligned with the downstream task, not universal
rules applied blindly.

---

# References

LLinás Solano, H. (2026). *Mathematics behind language representation: Lexical foundations and
vocabulary construction in NLP*. Department of Mathematics and Statistics, Universidad del Norte.
Retrieved from https://rpubs.com/hllinas/R_AI1_toc

Arlen, H., & Harburg, E. Y. (1939). *Somewhere over the rainbow* [Song]. From *The Wizard of Oz*.
Metro-Goldwyn-Mayer.
